---
title: "Statistical image processing: segmentation, phenotyping, and normalization"
output:
  html_document: 
    toc: true
    toc_float: true
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

```




## Slide Deck

<iframe class="speakerdeck-iframe" frameborder="0" src="https://speakerdeck.com/player/ccecd0548b874505b4bdd80172769ffe" title="MI_normalization" allowfullscreen="true" style="border: 0px; background: padding-box padding-box rgba(0, 0, 0, 0.1); margin: 0px; padding: 0px; border-radius: 6px; box-shadow: rgba(0, 0, 0, 0.2) 0px 5px 40px; width: 100%; height: auto; aspect-ratio: 560 / 314;" data-ratio="1.78343949044586"></iframe>


## Overview

This module performs and evaluates intensity normalization using the `mxnorm` package.

Load data and libraries.

```{r}
library(tidyverse)
library(viridis)
library(patchwork)

# load processed lung cancer data
load(url("https://github.com/julia-wrobel/MI_tutorial/raw/main/DataCode/lung.RDA"))

# for computational reasons, I will perform on a subset of the data with 20 subjects
ids = sample(unique(lung_df$patient_id), 20)
lung_df  = lung_df %>% filter(patient_id %in% ids)
```



## Normalization

I will use the `mxnorm` package to normalize the lung data. More examples are available in the package vignette.

### Setting up `mx_dataset` object

The `mxnorm` package uses an S3 object-oriented framework in R to define an mx_dataset which all operations are performed on.

* Use `patient_id` as slide identifier
* Use `image_id` as image identifier
* Use the following marker columns:
  * `cd19`
  * `cd14`
  * `cd3`
  * `cd8`
  * `hladr`
  * `ck`
  * `dapi`
* Use `tissue_category` as a metadata column


```{r}
# load the package
# install.packages("mxnorm")
library(mxnorm)

# define the mx_dataset
mx_data = mx_dataset(data = lung_df,
                     slide_id = "patient_id",
                     image_id = "image_id",
                     marker_cols = c("cd19",
                                     "cd3",
                                     "cd14",
                                     "cd8",
                                     "hladr",
                                     "ck",
                                     "dapi"
                                     ),
                     metadata_cols = c("tissue_category", 
                                       "phenotype_ck", 
                                       "phenotype_cd8", 
                                       "phenotype_cd14",
                                       "phenotype_other", 
                                       "phenotype_cd19"))


class(mx_data)
summary(mx_data)
```


And let's look at the data object we're going to use going forward:

```{r}
knitr::kable(head(mx_data$data))  %>%
    kableExtra::kable_styling() %>% kableExtra::scroll_box(width = "100%")
```


### Normalize using `mx_normalize()`

Here I normalize using the transformation: `log10mean_divide`. Then, I calculate Otsu discordance metrics.


```{r mxnorm}
mx_norm = mx_normalize(mx_data,
                       transform = "log10_mean_divide",
                       method = "None")


# Otsu
mx_norm = run_otsu_discordance(mx_norm,
                            table="both",
                            plot_out = FALSE)

```


Summary table that calculates the metrics from the Bioinformatics paper is printed below

```{r mxnorm_stats}
# calculate metrics for unnormalized and normalized data
summary(mx_norm)
```

### Plotting

First, density plots. Only showing density plots for unnormalized cells for two types of immune markers and 3 subjects.

```{r density_mxnorm}
set.seed(103001)
ids = sample(unique(lung_df$patient_id), 3)
markers = c("cd19", "cd14")

mx_df = mx_norm$data %>%
  mutate(cd14_norm = mx_norm$norm_data$cd14) %>%
  dplyr::select(patient_id, image_id, cd14, cd14_norm) %>%
  pivot_longer(cd14:cd14_norm, 
               names_to = "marker", 
               values_to = "marker_value") %>%
  filter(patient_id %in% ids)
  
  
mx_df %>% 
  ggplot() +
  geom_line(stat = "density", aes(marker_value, group = image_id, 
                                  color = patient_id), 
            alpha=0.5, linetype = 2) +
    geom_density(aes(marker_value, group = patient_id,
                   color = patient_id), size = 1.25) +
  scale_color_viridis(discrete = TRUE) +
  facet_wrap(~marker) +
  #theme(legend.position = "none") +
  xlim(0, 1) +
  labs(x = "")


```


Discordance plots. Show for subset of markers and maybe subset of subjects.

```{r discordance_mxnorm, eval = FALSE}
set.seed(103001)
ids = sample(unique(lung_df$patient_id), 15)

otsu_data = mx_norm$otsu_data %>%
  filter(marker %in% c("cd14", "cd19", "cd8", "cd3", "dapi", "ck"),
         slide_id %in% ids) %>%
  mutate(slide_id = as.numeric(factor(slide_id)),
         norm = factor(table, levels = c("raw", "normalized"))) 

point_size = 2
    

mean_vals = otsu_data %>% 
  group_by(norm, slide_id) %>% 
  summarize(m1 = mean(discordance_score), .groups = "drop")
        

otsu_data %>%
  ggplot() + 
  geom_point(aes(discordance_score, slide_id, color = marker), size = point_size) +
  facet_wrap(~ norm) +
  geom_point(data = mean_vals, aes(m1, slide_id), color = "black", fill = "white", 
             shape = 23, size = point_size) +
  labs(x = "discordance score", y = "slide_id")

```

```{r}
rm(mx_data, mx_norm, mean_vals, otsu_data, discordance_score, ids, markers, point_size, mx_df)
```



## References

* [textbook chapter, Wrobel and Vandekar](http://juliawrobel.com/Downloads/mIF_chapter.pdf)
* [Challenges and Opportunities in the Statistical Analysis of Multiplex Immunofluorescence Data](https://www.mdpi.com/2072-6694/13/12/3031)

### Segmentation

* [Mesmer deep learning based cell segmentation](https://pubmed.ncbi.nlm.nih.gov/34795433/)
* [Two-stage R-CNN](https://www.nature.com/articles/s41598-022-08355-1)
* [Cell Profiler](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2006-7-10-r100)
* [ilastik](https://www.ilastik.org/)
* [Halo (proprietary) software](https://indicalab.com/halo/)
* [inForm (proprietary) software](https://www.akoyabio.com/phenoimager/software/inform-tissue-finder/)


### Phenotyping

* [Astir](https://www.sciencedirect.com/science/article/pii/S2405471221003355)
* [Evaluation of clustering algorithms on Vectra and MIBI data](https://bmcresnotes.biomedcentral.com/articles/10.1186/s13104-022-06097-x)
* [MAUI phenotyping for MIBI images](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008887)
* [CU-Anschutz pixel-based platform independent phenotyping](https://www.biorxiv.org/content/10.1101/2022.10.20.510630v1.full.pdf)
* [CELESTA, nature methods](https://pubmed.ncbi.nlm.nih.gov/35654951/)
* [Gamma mixture models for phenotyping](https://arxiv.org/abs/2110.13074)

### Normalization


* [Python-based pipeline](https://www.nature.com/articles/s42003-022-03368-y)
* [mxnorm package](https://joss.theoj.org/papers/10.21105/joss.04180.pdf) 
* [Methods behind mxnorm](https://academic.oup.com/bioinformatics/article/38/6/1700/6496920)
